# Module 1: Bonus Exercise Results {-}

Load packages and data, subset needed columns:
```{r, class.source="codeblock",eval=TRUE}
library(MASS)
data(crabs)
```

Learn more about the data:
```{r, class.source="codeblock",eval=TRUE}
?crabs
head(crabs)
```

Subset needed columns:
```{r, class.source="codeblock",eval=TRUE}
crabs_meas <-  crabs[,c("FL","RW","CL","CW","BD")]
```

Perform hierarchical clustering:

```{r, class.source="codeblock",eval=TRUE}
c_dist <- dist(crabs_meas)
c_hclust <- hclust(c_dist)
plot(c_hclust)
```

Colour-code samples based on cluster assignment. Assume there are two clusters.

```{r, class.source="codeblock",eval=TRUE}
c_clusters = cutree(c_hclust,k = 2)
```

Now create a pairs plot, but colour-code by: 
1. by gene-expression based clusters
2. by species
3. by sex

```{r, class.source="codeblock",eval=TRUE}
pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[c_clusters]
)

pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[factor(crabs$sp)]
)

pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[factor(crabs$sex)]
)
```

Hierarchical clustering:
```{r, class.source="codeblock",eval=TRUE}
h <- hclust(dist(crabs_meas),method="ward.D2")
c2 <- cutree(h,k=2)

hclust_fun <- function(x){
    f <- hclust(x, method = "ward.D2");
    return(f)
}

library(RColorBrewer)
heatmap(
    as.matrix(crabs_meas),
    hclustfun = hclust_fun,
    col = brewer.pal("Blues",n=8),
    RowSideColors = c("pink","brown")[c2], 
    ColSideColors = rep("green",5)
)
```

Plot by sex:
```{r, class.source="codeblock",eval=TRUE}
heatmap(
    as.matrix(crabs_meas),
    hclustfun = hclust_fun,
    col = brewer.pal("Blues",n=8),
    RowSideColors = c("pink","brown")[factor(crabs$sex)], 
    ColSideColors = rep("green",5)
)
```


