# Module 3: Generalized Linear Models {-}

 In this module we're going to cover:

 * Reading data from files and evaluating missingness
 * Creating publication-quality plots with `ggplot2`
 * Fitting models with binary outcomes using generalized linear models

## Essential R: Reading tables from files, merging, basic data exploration  {-}

For this we're going to use two data files available in the course data directory.

[Download these](https://drive.google.com/drive/folders/1hKRa7ui-DmQCbqoKlgkmvhlrbXDmbgsV?usp=sharing) for use and put them in your R working directory.

* `tomerge1.csv` : comma-separated values
* `tomerge2.txt` : space-delimited values

Use `read.delim()` to read in tables. Note the use of the `sep=` parameter to indicate what the column separator is:

```{r, class.source="codeblock",eval=TRUE}
x1 <- read.delim("AUR2024_data/tomerge1.csv",sep=",")
head(x1)

x2 <- read.delim("AUR2024_data/tomerge2.txt",sep=" ")
head(x2)
```

Use `merge()` to combine the two tables by sample ID. Note the use of `by.x` and `by.y` to tell `merge()` which columns are equivalent:

```{r, class.source="codeblock", eval=TRUE}
x_merge <- merge(x1, x2, by.x = "Sample_ID", by.y = "sampleID")
head(x_merge)
```

We're going to use a popular dataset for data analysis, pertaining to survival of passengers aboard the Titanic. 

Download [the dataset here](https://drive.google.com/file/d/1PdP7bZTbmcRrPPPYFf1Mr8vfucom-d4k/view?usp=sharing) and copy it into your working directory.

Let's read in the data from a file:
```{r, class.source="codeblock",eval=TRUE}
dat <- read.delim(
    "AUR2024_data/titanic.csv",
    sep="," # indicate the column separator
    ) 
```

Examine the columns:
```{r, class.source="codeblock",eval=TRUE}
head(dat)
```

Some of the columns are categorical, use `table()` to look at the tallies:
Examine the columns:
```{r, class.source="codeblock",eval=TRUE}
table(dat$Survived)
table(dat$Pclass)
```

Use `summary()` to look at continuous-valued data:
```{r, class.source="codeblock",eval=TRUE}
summary(dat$Age)
```

Notice that there are 177 `NA` (missing) values for age.
Let's visualize the missing data more systematically.

## Explore missing data {-}

For this let's use a little script that converts a table into black and white squares to visualize missing data. For this, install the `plotrix` package

```{r, class.source="codeblock",eval=TRUE}
if (!requireNamespace("plotrix", quietly = TRUE)) install.packages("plotrix")

suppressMessages(library(plotrix))

#' show data missingness as a chequered matrix
#' 
#' @param x (matrix) data matrix.
#' @param outFile (char) path to file for printing graph
#' @param wd (numeric) width in inches
#' @param ht (numeric) height in inches
#' @return plots missingness matrix to file
#' @import plotrix
#' @export
plotMissMat <- function(x,xlab="columns",
		ylab="rows",border=NA) {
	
	x <- !is.na(x)
	class(x) <- "numeric"
	color2D.matplot(x,show.values=FALSE,axes=FALSE,
		cs1=c(1,0),cs2=c(1,0),cs3=c(1,0),border=border,
		cex=0.8,
		xlab=xlab,ylab=ylab)
}
```

Let's look at the missingness in the Titanic dataset. Missing data is shown as a white cell, and non-missing data is shown in black.

```{r, class.source="codeblock",eval=TRUE}
plotMissMat(dat)
```

We can see a column with many missing values. This is probably the "age" data. 
Let's count the number of missing values on a per-column level.

For this we combine `is.na()`, which returns a TRUE/FALSE value for NA values, and `colSums()` which adds up the TRUE values down the columns.

```{r count-missing, class.source="codeblock",eval=TRUE}
colSums(is.na(dat))
```
This confirms that `Age` is the only column with missing data.

Now let's explore the data using plots.

## Essential R: Plots with `ggplot2`  {-}

`ggplot2` is a popular plotting package that uses an additive paradigm to build plots. 
 
Useful websites:
* [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
* The [ggplot2 website](https://ggplot2.tidyverse.org/) is a wealth of reference information, so here we just touch on the basics to get you started. 


Anytime you need to generate a specific kind of plot, the website will most likely have documentation for how to achieve it.

Let's start by creating a scatterplot with two **continuous variables**. For this let's load a dataset measuring statistics around quality of life in US states in the late 70's:

```{r, class.source="codeblock",eval=TRUE}
state.x77 <- as.data.frame(state.x77)
head(state.x77)
```

Create a base plot using the `ggplot` function. Then "add" a scatterplot to it. Notice that the plot has been assigned to a variable named `p`. 

This setup is standard for `ggplot2` and allows multiple visualizations to be applied to the same base plot.

We use `aes` to tell ggplot what the `x` and `y` axes are, and later, if we want to colour-code by a particular column.

```{r load-states-data, class.source="codeblock",eval=TRUE}
library(ggplot2)
p <- ggplot(state.x77, 
    aes(x = Illiteracy,y = Income)
    )

p <- p + geom_point() # scatter plot
p
```

Now let's add confidence intervals:
```{r states-geomsmooth, class.source="codeblock",eval=TRUE}
p + geom_smooth()
```

It looks like there is a negative relationship between illiteracy and income.
We can confirm this by looking at correlation:
```{r test-cor, class.source="codeblock",eval=TRUE}
x <- state.x77$Illiteracy
y <- state.x77$Income
cor.test(x,y)
cor.test(x,y)$p.value
```

Let's now examine **categorical variables**. 
In the titanic set, let's look at the fare paid based on passenger class:
```{r boxplot, class.source="codeblock",eval=TRUE}
p <- ggplot(dat)
p + geom_boxplot(
    aes(x = factor(Pclass), # "factor()" makes a data column a categorical variable
        y = Fare))
```

We can use barplots to examine counts and proportions.
Let's look at number of survivors, split by passenger class.

Here the `fill` command is used to to split each barplot by the category, "Survived".
So you can see number of passengers by "Pclass" split by survival.

```{r  barplot, class.source="codeblock",eval=TRUE}
p + geom_bar(
    aes(fill=factor(Survived), 
        x = Pclass)
)
```

The plot above shows count data. Let's convert this to proportions. We can see that the fraction of non-survivors in "Class 3" is high.
```{r barplot2 , class.source="codeblock",eval=TRUE}
p + geom_bar(
    aes(fill=factor(Survived), x = Pclass),
    position = "fill"
)
```

How about males versus females?
```{r finish-titanic, class.source="codeblock",eval=TRUE}
p + geom_bar(
    aes(fill=factor(Survived), x = Sex),
    position = "fill"
)
```

**Exercise for later:** Try other `ggplot` functions on these data or on a small data table of your project.
(avoid using large genomics datasets, because those are going to be harder to interpret)

## Fit binary response variable using `glm()` and logistic regression {-}

Let's fit a model to a binary outcome. For this we load a dataset that measures physiological variables in a cohort of Pima Indians.

```{r load-pima, class.source="codeblock",eval=TRUE}
library(mlbench)
data(PimaIndiansDiabetes2)
# type ?PimaIndiansDiabetes2 to learn more about the dataset.

dat <- PimaIndiansDiabetes2
head(dat)
```

Let's look at the impact of blood glucose levels on diabetes diagnosis. 

First let's make a scatterplot.  Could there be a relationship?

```{r plot-diabetes-vs-glucose, class.source="codeblock",eval=TRUE}
p <- ggplot(dat, aes(x = glucose, y = factor(diabetes)))
p + geom_point()
```

Could this be fit with a linear model? 

Is there a continuous line that could reasonably fit the data?

```{r, class.source="codeblock",eval=TRUE}
p <- ggplot(dat, aes(x = glucose, y = factor(diabetes)))
p + geom_point() + geom_smooth()
```

This is a situation where a logistic regression model would be an appropriate choice because of the binary outcome. 

We're going to use `glm()` to fit a model to these data:

```{r diabetes-model , class.source="codeblock",eval=TRUE}
mod <- glm(factor(diabetes)~glucose, 
    dat,
    family = "binomial" # set to model binary outcome
)
summary(mod)
```

Which factors explain a diabetes diagnosis?

What if we include a couple other factors?
```{r diabetes-model-expanded, class.source="codeblock",eval=TRUE}
mod <- glm(factor(diabetes)~ glucose + pregnant + age + pressure + triceps,
    dat,
    family = "binomial")
summary(mod)
```

```{block,type="rmd-caution"}
Note: This morning's session only intends to introduce you to fitting non-linear models. 

In practice you may need to do more work to test multiple models to ascertain best fits your data, using measures such as goodness-of-fit. You will also likely compute odds ratio (odds of increased Y per unit increase X), which is out of scope for the current tutorial. 

We strongly recommend that you learn these topics before applying these methods to your own data.
```

## Bonus Exercise {-}

Now let's apply the ideas above to a dataset for classifying a breast cell as being either benign or malignant. 

```{r bc-load, class.source="codeblock",eval=TRUE}
data(BreastCancer)
bc <- BreastCancer
for (k in 2:10) # altered for current lab
    bc[,k] <- as.numeric(bc[,k]) 
head(bc)
```

Learn more about the dataset:
```{r bc-explore, class.source="codeblock",eval=TRUE}
?BreastCancer
```

For your exercise, answer the following questions:

* Is there missing data?
* Which columns are 
* Use plots to explore the relationship between explanatory variables.
* Fit a logistic model to identify which factors explain class (benign vs. malignant).